Intervention Bandits

### 2014-11-14 causal-bandit/c-bandit.tex
updates notation and adds latest plots to c-bandit.aux

\begin{document}

\maketitle

@@

\begin{enumerate}
\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
\eq{
\tilde X_{t,i} = \begin{cases}
X_{t,i} &\text{if } i \neq I_t \\
J_t & \text{otherwise}\,.
\end{cases}
}
\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
\end{enumerate}
The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|X_i = j]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
\eq{
R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
}

\section{Estimating $\mu_{i,j}$}
The most natural way to estimate $\mu_{i,j}$ is to compute an empirical estimate based on samples when that action was taken. This approach would
lead directly to the UCB algorithm with $2K$ actions and a regret bound that depended linearly on $K$. In this instance we can significantly
outperform this approach by exploiting the extra structure in the problem.

Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
We define estimator $\hat \mu_t$ of $\mu_{i,j}$ by
\eq{
m_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b, Y_t = 1} \\
n_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b} \\
\hat \mu_t &= \eta_i \frac{m_{i,j}}{n_{i,j}}  + \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right] \\
\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \\
n_a        &= \begin{cases} 
  n_{i,j}                                           & \text{if } a = i \\
  \frac{1}{2}\min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
\end{cases}
}
\begin{theorem} With probability at least $1 - \delta$ we have that:
$\displaystyle \left|\hat \mu_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
where $\beta > 0$ is some constant.
\end{theorem}

@@

@@

For small number of arms there is not much difference. As the number of arms increases, cusb shows significantly lower regret. As the number of arms increases above 100 the regret of the two algorithms begins to converge again. I suspect this is due to the fixed time horizon of $1000$ steps. With a time horizon of $100$ steps the two algorithms begin to converge again at a smaller number of arms.

@@

>>>>

\begin{document}
\def\ci{\perp\!\!\!\perp}
\maketitle

@@

\begin{enumerate}
\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
\eq{
\tilde X_{t,i} = \begin{cases}
X_{t,i} &\text{if } i \neq I_t \\
J_t & \text{otherwise}\,.
\end{cases}
}
\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
\end{enumerate}
The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|do(X_i = j)]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
\eq{
R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
}

\section{Estimating $\mu_{i,j}$}
The most natural way to estimate $\mu_{i,j}$ is to compute an empirical estimate based on samples when that action was taken. This approach would
lead directly to the UCB algorithm with $2K$ actions and a regret bound that depended linearly on $K$. In this instance we can significantly
outperform this approach by exploiting the known causal structure of the problem.

\eq {
P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
 &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
 &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
 &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b) 
}

Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.

Let $\hat{\mu_a}$ be an empirical estimator for $P(Y|do(X_i=j))$ obtained via marginalization over $a$.

\eq {
\hat{\mu_a} &=
\begin{cases}
\frac{m_{a,1}}{n_{a,1}}p_a+\frac{m_{a,0}}{n_{a,0}}(1-p_a) & \text{if } a \neq{i} \\
\frac{m_{i,j}}{n_{i,j}} & \text{if } a=i \\
\end{cases}\\
\text{where:}\\
m_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b, Y = 1}_s \\
n_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b}_s \\
}

This gives $K$ estimators $\{\hat{\mu_1}...\hat{\mu_K}\}$ to be pooled into a single estimator $\hat{\mu}$.

\eq {
\hat{\mu} &= \sum^{K}_{a=1}\eta_a\hat{\mu}_a=\eta_i \frac{m_{i,j}}{n_{i,j}}+ \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right]\\
\text{where:}\\
\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \text{ and } n_{i,j}=\begin{cases} 
n_{i,j} & \text{if } a = i \\
\frac{1}{2}\min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
\end{cases}
}

If $p$ is not known, these expression are unchanged except that $p_a$ is replaced with $\hat{p_a}$

\eq{
\hat{p_a} = \frac{\sum_{s=1}^{t} \ind{X_a=1,I \neq a}_s}{\sum_{s=1}^{t} \ind{I \neq a}_s}
}

\begin{theorem} With probability at least $1 - \delta$ we have that:
$\displaystyle \left|\hat{\mu}_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
where $\beta > 0$ is some constant.
\end{theorem}

@@

\begin{figure}[h]
\caption{Comparison of the performance of standard UCB versus causal UCB with $\beta=3$ and $\beta = 5$. 100 simulations were run for each algorithm up to a horizon of $10^5$ per value of $K$. Error bars span the 1st to 3rd quantile of the regret, round points mark the median and triangular points show the mean. For standard UCB the regret increases linearly with the number of arms $K$. For causal UCB the increase is sub-linear. Increasing $\beta$ leads to slower convergence but lower variance.}
\includegraphics[width=1\textwidth]{regret_vs_K_series_beta.pdf}
\end{figure}

\begin{figure}[h]
\caption{The distribution of regret varies with the $\beta$ parameter in the bound in the estimator. As beta increases, the mean regret increases but the variance decreases. The plot shows the results of running $100$ independent bandits, with $K=256$ and $\epsilon=0.1$, up to a horizon $h=10^5$ for each value of $\beta$. }
\includegraphics[width=0.7\textwidth]{regret_vs_beta_boxplot.pdf}
\end{figure}

@@

@@

\begin{figure}[h]
\caption{Regret vs number of arms with horizon $100$}
\label{fig:regretvsarmsshorthorizon}
\includegraphics[width=0.8\textwidth]{regregvsamrs100.pdf}
\end{figure}
